---
title: "Lab 2 - Data wrangling"
author: "Luke Owen"
date: "March 16, 2026"
format:  
  html: default # for speed
  pdf: default # for final output

---


```{r}
#| eval: true
#| message: false

library(tidyverse)
write_csv(midwest, "midwest.csv")
```


# Questions

## Part 1

### Question 1
Calculate the number of counties in each state and display your results in descending order of number of counties. Which state has the highest number of counties, and how many? Which state has the lowest number, and how many?
```{r}
midwest |>
  count(state, name = "n_counties") |>
  arrange(desc(n_counties))
```
Illinois has the highest number of counties, with 102. Wisconsin has the lowest number of counties, with 72.
\newpage

### Question 2
A classmate says “Look at that, there is a county called ___ in each state in this dataset!” In a single pipeline, discover all counties that could fill in the blank. Your response should be a data frame with only the county names that could fill in the blank and how many times they appear in the data.

```{r}
midwest |>
  count(county, name = "n") |>
  filter(n == 5)
```

There are three counties that share names across all states. Crawford, Jackson, and Monroe.
\newpage

### Question 3

3a.The counties with a population density higher than 25,000. Your code must use the filter() function.

```{r}
midwest |>
  select(county, state, popdensity, poptotal, area) |>
  filter(popdensity > 25000) |>
  arrange(desc(popdensity))
```

3b. The county with the highest population density. Your code must use the max() function.

```{r}
midwest |>
  select(county, state, popdensity, poptotal, area) |>
  filter(popdensity == max(popdensity))
```
\newpage

### Question 4

```{r}
midwest |>
  summarize(
    median_density = median(popdensity),
    q1_density = quantile(popdensity, 0.25),
    q3_density = quantile(popdensity, 0.75)
  )
```
The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has a population density of 1156 people per unit area. The middle 50% of counties have population densities between 622 and 2330 people per unit area.

\newpage

### Question 5

Another visualization from Lab 1 was the following that showed the proportion of urban counties in each state.Calculate the proportions in a single data pipeline.

```{r}
midwest |>
  mutate(metro = if_else(inmetro == 1, "Yes", "No")) |>
  group_by(state) |>
  summarize(proportion_metro = mean(metro == "Yes"))
```

\newpage

### Question 6

6.a. In a single pipeline, identify the observations marked in the orange square on the upper left corner. Your answer should be a data frame with four variables: county, state, percentage below poverty, and percentage college educated.

```{r}
midwest |>
  filter(percollege < 10, percbelowpoverty > 30) |>
  select(county, state, percbelowpoverty, percollege)
```

6.b. In a single pipeline, identify the observations marked in the red square in the plot above. Your answer should again be a data frame with four variables: county, state, percentage below poverty, and percentage college educated.

```{r}
midwest |>
  filter(percollege > 40, percbelowpoverty < 20) |>
  select(county, state, percbelowpoverty, percollege)
```

6.c. Bring your answers from part (a) and part (b) together! In a single pipeline, and a single filter() statement, identify the observations marked in the red and orange square in the plot above. Your answer should again be a data frame with four variables: county, state, percentage below poverty, and percentage college educated.

```{r}
midwest |>
  filter(
    (percollege < 10 & percbelowpoverty > 30) |
    (percollege > 40 & percbelowpoverty < 20)
  ) |>
  select(county, state, percbelowpoverty, percollege)
```

6.d. Create a new variable in midwest called potential_outlier. This variable should take on the value:

Yes if the point is one the ones you identified in part (c), i.e., one of the points marked in the squares in the plot above.

No otherwise.

Then, display the updated midwest data frame, with county, state, percentage below poverty, percentage college educated, potential_outlier as the selected variables, arranged in ascending order of potential_outlier.

```{r}
midwest <- midwest |>
  mutate(
    potential_outlier =
      if_else(
        (percollege < 10 & percbelowpoverty > 30) |
        (percollege > 40 & percbelowpoverty < 20),
        "Yes",
        "No"
      )
  )

midwest |>
  select(county, state, percbelowpoverty, percollege, potential_outlier) |>
  arrange(potential_outlier)
```

6.e. Recreate the visualization above, i.e. a scatterplot of percentage below poverty vs. percentage of people with a college degree, however color the points by the newly created potential_outlier variable instead of state.

```{r}
ggplot(midwest, aes(x = percollege, y = percbelowpoverty, color = potential_outlier)) +
  geom_point() +
  labs(
    x = "Percent with college degree",
    y = "Percent below poverty",
    title = "Potential outlier counties based on education and poverty"
  ) +
  theme_minimal()
```

\newpage

### Question 7

7.a. In a single pipeline, calculate the total population for each state and save the resulting data frame as state_population and display it in descending order of total population.

```{r}
state_population <- midwest |>
  group_by(state) |>
  summarize(total_population = sum(poptotal)) |>
  arrange(desc(total_population))

state_population
```

7.b. Then, in a separate pipeline, calculate the proportion of the total population in each state and, once again, display the results in descending order of proportion of population.

```{r}
state_population |>
  mutate(population_proportion = total_population / sum(total_population)) |>
  arrange(desc(population_proportion))
```

7.c. Which Midwestern state is most populous and what percent of the Midwest population lives there? Which is the least populous and what percent lives there?

The most populous Midwestern state is Illinois, and approximately 27% of the Midwest population lives there. The least populous state is Wisconsin, and approximately 12% of the Midwest population lives there.

\newpage

### Question 8



\newpage

## Part 2


### Question 9

